---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
if ("shared_home" %in% names(formals(duckdb::duckdb))) {
  Sys.setenv(DUCKDB_R_HOME = path.expand("~/.duckdb"))
}
```

# RBCFTools

<!-- badges: start -->
[![R-CMD-check](https://github.com/RGenomicsETL/RBCFTools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RGenomicsETL/RBCFTools/actions/workflows/R-CMD-check.yaml)
[![R-universe version](https://RGenomicsETL.r-universe.dev/RBCFTools/badges/version)](https://RGenomicsETL.r-universe.dev/RBCFTools)
<!-- badges: end -->

RBCFTools provides [FastDup](https://github.com/zzhofict/FastDup), [Samtools](https://github.com/samtools/samtools), [BCFtools](https://github.com/samtools/bcftools), and [HTSlib](https://github.com/samtools/htslib), the standard tools for reading and manipulating VCF/BCF and SAM/BAM/CRAM files. The package bundles these libraries and command-line tools (fastdup, samtools, bcftools, bgzip, tabix), so no external installation is required. All bundled tools are compiled against the same vendored HTSlib source and build. FastDup carries an explicit, tested RBCFTools patch overlay documented in [ERRATA.md](ERRATA.md). When compiled with libcurl, remote file access from S3, GCS, and HTTP URLs is supported. The package also includes support for streaming VCF/BCF to Apache Arrow (IPC) format via [nanoarrow](https://arrow.apache.org/nanoarrow/), with export to Parquet format using [duckdb](https://duckdb.org/) via either the [DuckDB nanoarrow extension](https://duckdb.org/community_extensions/extensions/nanoarrow.html) or a [`bcf_reader`](https://github.com/RGenomicsETL/RBCFTools/tree/v1.24-1.1.0/inst/duckdb_bcf_reader_extension) extension bundled in this package.

## Installation

You can install RBCFTools from r-universe on Unix-alike systems. Windows is not supported.
 
```r
install.packages(
    'RBCFTools',
    repos = c(
              'https://rgenomicsetl.r-universe.dev',
              'https://cloud.r-project.org')
      )
```

## Version Information

```{r versions}
library(RBCFTools)
# Get bundled software versions
fastdup_version()
samtools_version()
bcftools_version()
htslib_version()
```

## Tool Paths

RBCFTools bundles FastDup, Samtools, BCFtools, and HTSlib command-line tools. Use the path functions to locate the executables.

```{r paths}
fastdup_path()
samtools_path()
bcftools_path()
bgzip_path()
tabix_path()
# List all available tools
fastdup_tools()
samtools_tools()
bcftools_tools()
htslib_tools()
```

## Generic executable pipelines

`pipeline_stage()` and `run_pipeline()` connect arbitrary executables without constructing a shell command. Arguments are passed directly to each process, every stage's status is returned, and binary data moves through operating-system pipes rather than through R strings. On Linux, `cpu_affinity` starts every concurrent stage under the same `taskset` CPU set, while the result records native wall time, per-stage peak RSS, and the largest simultaneous sum of stage RSS sampled from `/proc` every 10 ms.

```{r pipeline-basic}
pipeline_output <- tempfile()
pipeline_status <- run_pipeline(
  list(
    pipeline_stage("printf", c("alpha\\nbeta\\n")),
    pipeline_stage("grep", "beta")
  ),
  stdout = pipeline_output
)
readLines(pipeline_output)
pipeline_status
```

A BWA or Rminibwa mapper that emits SAM can feed Samtools directly. Samtools duplicate marking requires name sorting and `fixmate -m` before coordinate sorting; omitting those steps does not produce a valid `samtools markdup` input. These examples are executed when the README is rendered. They build a deterministic reference, map real paired FASTQ records with both executables, mark duplicates, create indexes, and query the resulting BAM files. Rendering therefore requires BWA and Rminibwa rather than silently substituting a mock mapper.

```{r pipeline-alignment-fixture}
stopifnot(nzchar(Sys.which("bwa")), requireNamespace("Rminibwa", quietly = TRUE))
threads <- "2"
pipeline_dir <- tempfile("rbcftools-pipeline-")
dir.create(pipeline_dir)
reference_fasta <- file.path(pipeline_dir, "reference.fa")
reads_1 <- file.path(pipeline_dir, "reads_1.fastq")
reads_2 <- file.path(pipeline_dir, "reads_2.fastq")

set.seed(240814L)
reference_sequence <- paste0(sample(c("A", "C", "G", "T"), 50000L, replace = TRUE), collapse = "")
starts <- rep(501L + seq.int(0L, 19L) * 2000L, each = 2L)
inserts <- rep(300L + seq.int(0L, 19L) %% 5L * 10L, each = 2L)
read_1 <- substring(reference_sequence, starts, starts + 99L)
reverse_complement <- function(x) {
  paste0(rev(strsplit(chartr("ACGT", "TGCA", x), "", fixed = TRUE)[[1L]]), collapse = "")
}
read_2 <- vapply(
  seq_along(starts),
  function(i) reverse_complement(substring(
    reference_sequence,
    starts[[i]] + inserts[[i]] - 100L,
    starts[[i]] + inserts[[i]] - 1L
  )),
  character(1L)
)
read_names <- sprintf("INST:1:FC:1:1101:%d:%d", 100L + seq_along(starts), 200L + seq_along(starts))
write_fastq <- function(path, names, reads, mate) {
  records <- unlist(Map(
    function(name, read) c(paste0("@", name, "/", mate), read, "+", strrep("I", nchar(read))),
    names,
    reads
  ), use.names = FALSE)
  writeLines(records, path, useBytes = TRUE)
}
writeLines(c(">chr1", reference_sequence), reference_fasta, useBytes = TRUE)
write_fastq(reads_1, read_names, read_1, 1L)
write_fastq(reads_2, read_names, read_2, 2L)
data.frame(reference_bases = nchar(reference_sequence), read_pairs = length(read_names))
```

The BWA stream exercises the full Samtools-compatible duplicate-marking chain and then indexes the output:

```{r pipeline-samtools-markdup}
bwa_log <- file.path(pipeline_dir, "bwa.log")
bwa_output <- file.path(pipeline_dir, "bwa.samtools-markdup.bam")
run_pipeline(
  pipeline_stage("bwa", c("index", reference_fasta), name = "bwa index"),
  stderr = bwa_log
)
bwa_status <- run_pipeline(
  list(
    pipeline_stage(
      "bwa",
      c("mem", "-t", threads, reference_fasta, reads_1, reads_2),
      name = "bwa mem"
    ),
    pipeline_stage(samtools_path(), c("sort", "-n", "-@", threads, "-O", "BAM", "-o", "-", "-"), name = "name sort"),
    pipeline_stage(samtools_path(), c("fixmate", "-m", "-@", threads, "-", "-"), name = "fixmate"),
    pipeline_stage(samtools_path(), c("sort", "-@", threads, "-O", "BAM", "-o", "-", "-"), name = "coordinate sort"),
    pipeline_stage(samtools_path(), c("markdup", "-@", threads, "-", bwa_output), name = "markdup")
  ),
  stderr = bwa_log
)
bwa_index_status <- run_pipeline(
  pipeline_stage(samtools_path(), c("index", "-@", threads, bwa_output), name = "index")
)
bwa_status
bwa_index_status
data.frame(
  alignments = as.integer(system2(samtools_path(), c("view", "-c", bwa_output), stdout = TRUE)),
  duplicate_alignments = as.integer(system2(samtools_path(), c("view", "-c", "-f", "1024", bwa_output), stdout = TRUE)),
  bam_bytes = file.info(bwa_output)$size,
  index_exists = file.exists(paste0(bwa_output, ".bai"))
)
```

FastDup rereads a coordinate-sorted input and therefore forms an explicit file-dependent step rather than a streaming stage. Here the mapper is the actual executable exported by Rminibwa:

```{r pipeline-fastdup}
minibwa_log <- file.path(pipeline_dir, "minibwa.log")
minibwa_index <- file.path(pipeline_dir, "minibwa-index")
minibwa_sorted <- file.path(pipeline_dir, "minibwa.sorted.bam")
fastdup_output <- file.path(pipeline_dir, "minibwa.fastdup.bam")
run_pipeline(
  pipeline_stage(
    Rminibwa::minibwa_path(),
    c("index", "-t", threads, reference_fasta, minibwa_index),
    name = "minibwa index"
  ),
  stderr = minibwa_log
)
minibwa_status <- run_pipeline(
  list(
    pipeline_stage(
      Rminibwa::minibwa_path(),
      c("map", "-t", threads, minibwa_index, reads_1, reads_2),
      name = "minibwa map"
    ),
    pipeline_stage(
      samtools_path(),
      c("sort", "-@", threads, "-o", minibwa_sorted, "-"),
      name = "coordinate sort"
    )
  ),
  stderr = minibwa_log
)
fastdup_status <- run_pipeline(
  pipeline_stage(
    fastdup_path(),
    c(
      "--input", minibwa_sorted,
      "--output", fastdup_output,
      "--metrics", paste0(fastdup_output, ".metrics"),
      "--num-threads", threads,
      "--tagging-policy", "All"
    ),
    name = "fastdup"
  ),
  stderr = paste0(fastdup_output, ".log")
)
fastdup_index_status <- run_pipeline(
  pipeline_stage(samtools_path(), c("index", "-@", threads, fastdup_output), name = "index")
)
minibwa_status
fastdup_status
fastdup_index_status
data.frame(
  alignments = as.integer(system2(samtools_path(), c("view", "-c", fastdup_output), stdout = TRUE)),
  duplicate_alignments = as.integer(system2(samtools_path(), c("view", "-c", "-f", "1024", fastdup_output), stdout = TRUE)),
  bam_bytes = file.info(fastdup_output)$size,
  index_exists = file.exists(paste0(fastdup_output, ".bai"))
)
```

The rendered benchmarks execute the same pipelines on a [complete 10,110,535-pair HG02088 exome](https://github.com/RGenomicsETL/RBCFTools/blob/v1.24-1.1.0/benchmarks/hg02088-exome-pipelines/README.md) and on the pinned [Zenodo HG002 one-million-pair WGS dataset](https://github.com/RGenomicsETL/RBCFTools/blob/v1.24-1.1.0/benchmarks/hg002-wgs-1m-pipelines/README.md). Both map against a full human reference with BWA and Rminibwa, then compare Samtools `markdup` with the patched FastDup on each prepared BAM.

## Capabilities

Check which features were compiled into htslib

```{r capabilities}
# Get all capabilities as a named list
htslib_capabilities()

# Human-readable feature string
htslib_feature_string()
```



### Feature Constants

Use `HTS_FEATURE_*` constants to check for specific features
 
```{r features}
# Check individual features
htslib_has_feature(HTS_FEATURE_LIBCURL)  # Remote file access via libcurl
htslib_has_feature(HTS_FEATURE_S3)       
htslib_has_feature(HTS_FEATURE_GCS)      # Google Cloud Storage
htslib_has_feature(HTS_FEATURE_LIBDEFLATE)
htslib_has_feature(HTS_FEATURE_LZMA)
htslib_has_feature(HTS_FEATURE_BZIP2)
```

These are useful for conditionally enabling features in your code.

## Example Query Of Remote VCF from S3 using bcftools

With libcurl support, bcftools can directly query remote files. Here we count variants in a small region from the 1000 Genomes cohort VCF on S3:

```{r s3 setup, eval=TRUE}
# Setup environment for remote file access (S3/GCS)
setup_hts_env()

# Build S3 URL for 1000 Genomes cohort VCF
s3_base <- "s3://1000genomes-dragen-v3.7.6/data/cohorts/"
s3_path <- "gvcf-genotyper-dragen-3.7.6/hg19/3202-samples-cohort/"
s3_vcf_file <- "3202_samples_cohort_gg_chr22.vcf.gz"
s3_vcf_uri <- paste0(s3_base, s3_path, s3_vcf_file)
```

```{r s3-example, eval=TRUE}
# Use processx instead of system2
res <- processx::run(
  bcftools_path(),
  c("index", "-n", s3_vcf_uri),
  error_on_status = FALSE,
  echo = FALSE
)
res$stdou
```

##  VCF to Arrow Streams and Duckdb `bcf_reader` extension



RBCFTools provides streaming VCF/BCF to Apache Arrow Stream conversion via [nanoarrow](https://arrow.apache.org/nanoarrow/). This enables integration with tools like [DuckDB](https://github.com/duckdb/duckdb-r) and Parquet format conversion when serializing to Arrow IPC. We also support the [`bcf_reader`](https://github.com/RGenomicsETL/RBCFTools/tree/v1.24-1.1.0/inst/duckdb_bcf_reader_extension) DuckDB extension to read directly into DuckDB.

The `nanoarrow` stream conversion and `bcf_reader` perform VCF spec conformance checks on headers (similar to htslib's `bcf_hdr_check_sanity()`) and emit R warnings or DuckDB logs when correcting non-conformant fields


### Read VCF as Arrow Stream

Open BCF as Arrow array stream and read the batches


```{r arrow-stream, eval=TRUE}

bcf_file <- system.file("extdata", "1000G_3samples.bcf", package = "RBCFTools")
stream <- vcf_open_arrow(bcf_file, batch_size = 100L)

batch <- stream$get_next()

head(nanoarrow::convert_array(batch))
stream$release()
```

### Convert to Data Frame

Convert entire BCF to data.frame

```{r arrow-df, eval=TRUE}
options(warn = -1)  # Suppress warnings
df <- vcf_to_arrow(bcf_file, as = "data.frame")
df[, c("CHROM", "POS", "REF", "ALT", "QUAL")] |> head()

```


### Write to Arrow IPC

Arrow IPC (`.arrows`) format for interoperability with other Arrow tools using nanoarrow's native streaming writer 

```{r arrow-ipc, eval=TRUE}

# Convert BCF to Arrow IPC
ipc_file <- tempfile(fileext = ".arrows")

vcf_to_arrow_ipc(bcf_file, ipc_file)

# Read back with nanoarrow
ipc_data <- as.data.frame(nanoarrow::read_nanoarrow(ipc_file))
ipc_data[, c("CHROM", "POS", "REF", "ALT")] |> head()
```

### Write VCF Streams to Parquet

Using [DuckDB](https://github.com/duckdb/duckdb-r) to convert BCF to Parquet file and perform queries on the Parquet file. This involves VCF stream conversion to data.frame

```{r parquet, eval=TRUE}

parquet_file <- tempfile(fileext = ".parquet")
vcf_to_parquet_arrow(bcf_file, parquet_file, compression = "snappy")
con <- duckdb::dbConnect(duckdb::duckdb())
pq_bcf <- DBI::dbGetQuery(con, sprintf("SELECT * FROM '%s' LIMIT 100", parquet_file))
pq_me <- DBI::dbGetQuery(
    con, 
    sprintf("SELECT * FROM parquet_metadata('%s')",
    parquet_file))
duckdb::dbDisconnect(con, shutdown = TRUE)
pq_bcf[, c("CHROM", "POS", "REF", "ALT")] |>
  head()
pq_me |> head()
```

Use `streaming = TRUE` to avoid loading the entire VCF into R memory. This streams VCF to Arrow IPC (nanoarrow) and then to Parquet via DuckDB, trading memory for serialization overhead using the [DuckDB nanoarrow extension](https://duckdb.org/community_extensions/extensions/nanoarrow.html)

```{r streaming, eval = TRUE}
bcf_larger <- system.file("extdata", "1000G.ALL.2of4intersection.20100804.genotypes.bcf", package = "RBCFTools")
outfile <-  tempfile(fileext = ".parquet")
vcf_to_parquet_arrow(
    bcf_larger,
    outfile,
    streaming = TRUE,
    batch_size = 10000L,
    row_group_size = 100000L,
    compression = "zstd"
)

```

### Query VCF with duckdb after converting the Stream

SQL queries on BCF using DuckDB package. For now this is somewhat limited due to conversion from Arrow streams to data frame

```{r duckdb, eval=TRUE}
vcf_query_arrow(bcf_file, "SELECT CHROM, COUNT(*) as n FROM vcf GROUP BY CHROM")

# Filter variants by position
vcf_query_arrow(bcf_file, "SELECT CHROM, POS, REF, ALT FROM vcf  LIMIT 5")
```

###  Query VCF with  DuckDB Extension

A native DuckDB extension (`bcf_reader`) for direct SQL queries on VCF/BCF files without Arrow conversion overhead. The extension uses the DuckDB C API and should be compatible with DuckDB v1.2.0+

```{r duckdb-ext, eval=TRUE}
# Build extension (uses package's bundled htslib)
build_dir <- file.path(tempdir(), "bcf_reader")
ext_path <- bcf_reader_build(build_dir, verbose = FALSE)

# Connect and query a VCF.gz file
con <- vcf_duckdb_connect(ext_path)
vcf_file <- system.file("extdata", "test_deep_variant.vcf.gz", package = "RBCFTools")

# Describe schema
DBI::dbGetQuery(con, sprintf("DESCRIBE SELECT * FROM bcf_read('%s')", vcf_file))

# Aggregate query
DBI::dbGetQuery(con, sprintf("
  SELECT CHROM, COUNT(*) as n_variants, 
         MIN(POS) as min_pos, MAX(POS) as max_pos
  FROM bcf_read('%s') 
  GROUP BY CHROM 
  ORDER BY n_variants DESC 
  LIMIT 5", vcf_file))

# Export directly to Parquet
parquet_out <- tempfile(fileext = ".parquet")
DBI::dbExecute(con, sprintf("
  COPY (SELECT * FROM bcf_read('%s')) 
  TO '%s' (FORMAT PARQUET, COMPRESSION ZSTD)", vcf_file, parquet_out))

# Verify parquet file size
file.info(parquet_out)$size

# Same query on Parquet file
DBI::dbGetQuery(con, sprintf("
  SELECT CHROM, COUNT(*) as n_variants, 
         MIN(POS) as min_pos, MAX(POS) as max_pos
  FROM '%s' 
  GROUP BY CHROM 
  ORDER BY n_variants DESC 
  LIMIT 5", parquet_out))

DBI::dbDisconnect(con)
```

### Open VCF as DuckDB Table/View

The `vcf_open_duckdb()` function provides a convenient interface to open VCF/BCF files as DuckDB tables or views, similar to `vcf_open_arrow()` for Arrow streams. By default it creates a **lazy view** that reads from the VCF on each query - instant to create with no memory overhead.

```{r duckdb-open, eval=TRUE}
# Build extension
ext_path <- bcf_reader_build(tempdir(), verbose = FALSE)
vcf_file <- system.file("extdata", "1000G_3samples.vcf.gz", package = "RBCFTools")

# Open as lazy view (default) - instant creation
vcf <- vcf_open_duckdb(vcf_file, ext_path)
print(vcf)

# Query the view
DBI::dbGetQuery(vcf$con, "SELECT CHROM, POS, REF, ALT FROM variants LIMIT 5")

# Close connection
vcf_close_duckdb(vcf)
```

For repeated queries on the same data, materialize to a table with `as_view = FALSE`:

```{r duckdb-table, eval=TRUE}
# Materialize to in-memory table (slower to create, fast queries)
vcf <- vcf_open_duckdb(vcf_file, ext_path, as_view = FALSE)
print(vcf)

# Multiple fast queries on materialized data
DBI::dbGetQuery(vcf$con, "SELECT COUNT(*) as n FROM variants")
DBI::dbGetQuery(vcf$con, "SELECT CHROM, COUNT(*) as n FROM variants GROUP BY CHROM")

vcf_close_duckdb(vcf)
```

For large VCF files, use tidy format and parallel loading:

```{r duckdb-tidy-open, eval=TRUE}
# Tidy format with column selection
vcf <- vcf_open_duckdb(
  vcf_file, ext_path,
  tidy_format = TRUE,
  columns = c("CHROM", "POS", "REF", "ALT", "SAMPLE_ID", "FORMAT_GT")
)

DBI::dbGetQuery(vcf$con, "SELECT * FROM variants LIMIT 6")
vcf_close_duckdb(vcf)
```

### Tidy (Long) Format Export

Export VCF/BCF to a "tidy" format where each row is one variant-sample combination. The native `tidy_format` parameter in the bcf_reader extension transforms wide `FORMAT_<field>_<sample>` columns into a single `SAMPLE_ID` column plus `FORMAT_<field>` columns - much faster than SQL-level UNNEST.

```{r tidy-format, eval=TRUE}
# Build extension
ext_path <- bcf_reader_build(tempdir(), verbose = FALSE)

# Multi-sample VCF: 3 samples x 11 variants = 33 rows
vcf_3samples <- system.file("extdata", "1000G_3samples.vcf.gz", package = "RBCFTools")
tidy_out <- tempfile(fileext = ".parquet")

# Use tidy_format parameter directly
vcf_to_parquet_duckdb(vcf_3samples, tidy_out, extension_path = ext_path, tidy_format = TRUE)

# Query the tidy output
con <- duckdb::dbConnect(duckdb::duckdb())
DBI::dbGetQuery(con, sprintf("
  SELECT CHROM, POS, REF, SAMPLE_ID, FORMAT_GT 
  FROM '%s' 
  LIMIT 9", tidy_out))

# Verify row expansion: variants * samples
DBI::dbGetQuery(con, sprintf("
  SELECT COUNT(*) as total_rows, 
         COUNT(DISTINCT SAMPLE_ID) as n_samples
  FROM '%s'", tidy_out))

DBI::dbDisconnect(con, shutdown = TRUE)
```

This format is ideal for combining single-sample VCFs into cohort tables or appending to DuckLake tables with MERGE operations.

### VCF Header Metadata in Parquet

When exporting VCF to Parquet, the full VCF header is embedded as Parquet key-value metadata by default. This preserves all schema information (INFO, FORMAT, FILTER definitions, contigs, samples) enabling round-trip back to VCF format.

```{r parquet-metadata, eval=TRUE}
# Build extension
ext_path <- bcf_reader_build(tempdir(), verbose = FALSE)
vcf_file <- system.file("extdata", "1000G_3samples.vcf.gz", package = "RBCFTools")

# Export with embedded VCF header (default)
parquet_out <- tempfile(fileext = ".parquet")
vcf_to_parquet_duckdb(vcf_file, parquet_out, ext_path)

# Read back the metadata
meta <- parquet_kv_metadata(parquet_out)
print(meta)

# Extract the VCF header (stored with escaped newlines)
vcf_header <- meta[meta$key == "vcf_header", "value"]
substr(vcf_header, 1, 200)
```


### Stream Remote VCF using Arrow or DuckDB Extension

Stream remote VCF region directly from S3 using either `nanoarrow` based `vcf_stream` or `duckb` extension

```{r url-arrow, eval=TRUE}
s3_vcf_uri <- paste0(s3_base, s3_path, s3_vcf_file)

# Arrow stream
stream <- vcf_open_arrow(
    s3_vcf_uri,
    region = "chr22:16050000-16050500",
    batch_size = 1000L
)
df <- as.data.frame(nanoarrow::convert_array_stream(stream))
df[, c("CHROM", "POS", "REF", "ALT")] |> head()


# Query remote VCF with bcf_reader extension

vcf_query_duckdb(
    s3_vcf_uri,
    ext_path,
    region = "chr22:16050000-16050500",
    query = "SELECT CHROM, POS, REF, ALT FROM vcf LIMIT 5"
)


```
### Custom Index File Path

For region queries, an index file is required. By default, RBCFTools looks for `.tbi` (tabix) or `.csi` indexes using standard naming conventions. For non-standard index locations such as presigned URLs or custom paths, use the `index` parameter. Index files are only required for region queries; when reading an entire file without a region filter, no index is needed. VCF files try `.tbi` first then fall back to `.csi`, while BCF files use `.csi` only.

```{r index-param, eval=TRUE}
# Explicit index file path
bcf_file <- system.file("extdata", "1000G_3samples.bcf", package = "RBCFTools")
csi_index <- system.file("extdata", "1000G_3samples.bcf.csi", package = "RBCFTools")

stream <- vcf_open_arrow(bcf_file, index = csi_index)
batch <- stream$get_next()
nanoarrow::convert_array(batch)[, c("CHROM", "POS", "REF")] |> head(3)
# Alternative: htslib ##idx## syntax in filename
filename_with_idx <- paste0(bcf_file, "##idx##", csi_index)
stream2 <- vcf_open_arrow(filename_with_idx)
batch2 <- stream2$get_next()
nanoarrow::convert_array(batch2)[, c("CHROM", "POS", "REF")] |> head(3)
```

## Example Application: DuckLake ETL

[`DuckLake`](https://ducklake.select/) is an integrated data lake and catalog format. It has two components: a Parquet file storage and a metadata database layer.

### DuckLake ETL to MinIO Storage backend

To mimic an S3 backend for [`DuckLake`](https://ducklake.select/) storage we use [MinIO](https://github.com/minio/minio), we convert VCF files to Parquet and insert them into DuckLake and query the lake. The lake has a local DuckDB metadata database in this example


#### Setup up `minio` storage backend
We start by seting up and configuring a `minio` server

```{r ducklake-minio, eval=TRUE}
# DuckLake with S3-compatible storage using local MinIO
bin_dir <- file.path(tempdir(), "ducklake_bins")
dir.create(bin_dir, recursive = TRUE, showWarnings = FALSE)

# Use installed MinIO and MC binaries
minio_bin <- Sys.which('minio')
mc_bin <- Sys.which('mc')

# Start MinIO (ephemeral) and configure mc
stopifnot(nzchar(minio_bin), nzchar(mc_bin))
data_dir <- file.path(tempdir(), "ducklake_minio")
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
port_text <- Sys.getenv("RBCFTOOLS_README_MINIO_PORT")
if (!nzchar(port_text)) {
  python_bin <- Sys.which("python3")
  stopifnot(nzchar(python_bin))
  port_text <- trimws(processx::run(
    python_bin,
    c("-c", paste0(
      "import socket; s=socket.socket(); s.bind(('127.0.0.1',0)); ",
      "print(s.getsockname()[1]); s.close()"
    ))
  )$stdout)
}
port <- as.integer(port_text)
stopifnot(!is.na(port), port > 0L, port <= 65535L)
endpoint <- sprintf("127.0.0.1:%d", port)

# Start MinIO as an owned child process; processx also cleans it up if rendering
# exits before the explicit cleanup chunk.
minio_log <- tempfile("rbcftools-minio-", fileext = ".log")
minio_process <- processx::process$new(
  minio_bin,
  c("server", data_dir, "--address", endpoint),
  stdout = minio_log,
  stderr = "2>&1",
  cleanup = TRUE
)
pid <- minio_process$get_pid()
pid

# Poll readiness rather than relying on a fixed sleep.
mc_cmd_args <- c("alias", "set", "ducklake_local",
                 paste0("http://", endpoint), "minioadmin", "minioadmin")
mc_ready <- NULL
for (attempt in seq_len(100L)) {
  mc_ready <- processx::run(
    mc_bin, mc_cmd_args, error_on_status = FALSE, echo = FALSE
  )
  if (identical(mc_ready$status, 0L)) break
  if (!minio_process$is_alive()) {
    stop("MinIO exited before becoming ready; see ", minio_log)
  }
  Sys.sleep(0.1)
}
stopifnot(identical(mc_ready$status, 0L))

# Create bucket with unique name
bucket <- sprintf("readme-demo-%d", as.integer(Sys.time()))
bucket_cmd_args <- c("mb", paste0("ducklake_local/", bucket))
processx::run(mc_bin, bucket_cmd_args, echo = FALSE)

# Store variables for later use
minio_endpoint <- endpoint
bucket_name <- bucket
data_root_s3 <- paste0("s3://", bucket_name)
data_path_s3 <- paste0(data_root_s3, "/data/")
mc_path <- mc_bin
```

####  Attach a DuckLake

```{r ducklake-attach, eval=TRUE}
con <- duckdb::dbConnect(duckdb::duckdb(config = list(
  allow_unsigned_extensions = "true",
  enable_external_access = "true"
)))
stopifnot(DBI::dbIsValid(con))

DBI::dbExecute(con, "INSTALL httpfs")
DBI::dbExecute(con, "LOAD httpfs")
DBI::dbExecute(con, "INSTALL ducklake")
DBI::dbExecute(con, "LOAD ducklake")

# Provide S3 credentials/endpoints for MinIO
Sys.setenv(
  AWS_ACCESS_KEY_ID = "minioadmin",
  AWS_SECRET_ACCESS_KEY = "minioadmin",
  AWS_DEFAULT_REGION = "us-east-1",
  AWS_S3_ENDPOINT = paste0("http://", minio_endpoint),
  AWS_S3_USE_HTTPS = "FALSE",
  AWS_S3_FORCE_PATH_STYLE = "TRUE",
  AWS_EC2_METADATA_DISABLED = "TRUE"
)

ducklake_create_s3_secret(
  con,
  name = "ducklake_minio",
  key_id = "minioadmin",
  secret = "minioadmin",
  endpoint = minio_endpoint,
  region = "us-east-1",
  use_ssl = FALSE
)

metadata_path <- file.path(tempdir(), sprintf("ducklake_meta_%s.ducklake", bucket_name))
ducklake_attach(
  con,
  metadata_path = metadata_path,
  data_path = data_path_s3,
  alias = "lake",
  extra_options = list(OVERRIDE_DATA_PATH = TRUE)
)

# Work inside the attached lake database
DBI::dbExecute(con, "USE lake")
```

### Write a VCF into minio and register to the Lake
```{r ducklake-write, eval=TRUE}
# Ensure we are operating inside the DuckLake catalog
DBI::dbExecute(con, "USE lake")

# Load variants via fast VCF to Parquet conversion
vcf_file <- system.file("extdata", "test_deep_variant.vcf.gz", package = "RBCFTools")
ext_path <- bcf_reader_build(tempdir())
ducklake_load_vcf(
  con,
  table = "variants",
  vcf_path = vcf_file,
  extension_path = ext_path,
  threads = 1,
  tidy_format = TRUE
)
```

### Add another VCF file with different schema

DuckLake supports schema evolution via `ALTER TABLE`. Use `allow_evolution = TRUE` to automatically add new columns from incoming files.

```{r ducklake-add-second, eval=TRUE}
DBI::dbExecute(con, "USE lake")
variants_count <- DBI::dbGetQuery(con, "SELECT COUNT(*) as n FROM variants")$n[1]
variants_count

vcf_file2 <- system.file("extdata", "test_vep.vcf", package = "RBCFTools")
local_parquet2 <- tempfile(fileext = ".parquet")
vcf_to_parquet_duckdb(vcf_file2, local_parquet2, extension_path = ext_path, tidy_format = TRUE)

DBI::dbGetQuery(con, sprintf("SELECT COUNT(*) as n FROM read_parquet('%s')", local_parquet2))

s3_parquet2 <- paste0(data_path_s3, "variants/variants_vep.parquet")
mc_cmd_args <- c("cp", local_parquet2, paste0("ducklake_local/", bucket_name, "/data/variants/variants_vep.parquet"))
processx::run(mc_bin, mc_cmd_args, echo = FALSE)

DBI::dbGetQuery(con, sprintf("SELECT COUNT(*) as n FROM read_parquet('%s')", s3_parquet2))

ducklake_register_parquet(
  con,
  table = "lake.variants",
  parquet_files = s3_parquet2,
  create_table = FALSE,
  allow_evolution = TRUE
)

variants_count_after <- DBI::dbGetQuery(con, "SELECT COUNT(*) as n FROM variants")$n[1]
variants_count_after
```

### List lake content
```{r check-lake, eval=TRUE}
DBI::dbExecute(con, "USE lake")

DBI::dbGetQuery(con, "
  SELECT 
    COUNT(*) as total_variants,
    COUNT(DISTINCT CHROM) as n_chromosomes,
    MIN(POS) as min_position,
    MAX(POS) as max_position
  FROM variants
")

DBI::dbGetQuery(con, "DESCRIBE variants") |>
  head(20)
```

```{r ducklake-files, eval=TRUE}
ducklake_list_files(con, "lake", "variants")
```

### Snapshots and time travel

```{r ducklake-snapshots, eval=TRUE}
ducklake_snapshots(con, "lake") |> head()
ducklake_snapshots(con, "lake") |> tail()

ducklake_current_snapshot(con, "lake")
```

### Configuration

```{r ducklake-config, eval=TRUE}
ducklake_options(con, "lake")
ducklake_set_option(con, "lake", "parquet_compression", "zstd")
ducklake_options(con, "lake")
```

```{r ducklake-cleanup, eval=TRUE}
DBI::dbDisconnect(con, shutdown = TRUE)
minio_process$kill()
minio_process$wait()
```

### Supported Metadata Databases

DuckLake supports multiple catalog backends

- `DuckDB` : `ducklake:path/to/catalog.ducklake`
- `SQLite`: `ducklake:sqlite:path/to/catalog.db`  
- `PostgreSQL` : `ducklake:postgresql://user:pass@host:5432/db`
- `MySQL`: `ducklake:mysql://user:pass@host:3306/db`

#### Connection Methods

The package provides helpers for different metadata databases

#### Direct Connection

```r
# DuckDB backend
ducklake_connect_catalog(
  con,
  backend = "duckdb",
  connection_string = "catalog.ducklake",
  data_path = "s3://bucket/data/",
  alias = "lake"
)

# PostgreSQL backend
ducklake_connect_catalog(
  con,
  backend = "postgresql", 
  connection_string = "user:pass@host:5432/db",
  data_path = "s3://bucket/data/",
  alias = "lake"
)
```

##### Secret-based Connection:**
```r
# Create catalog secret
ducklake_create_catalog_secret(
  con,
  name = "pg_catalog",
  backend = "postgresql",
  connection_string = "user:pass@host:5432/db",
  data_path = "s3://bucket/data/"
)

# Connect using secret
ducklake_connect_catalog(
  con,
  secret_name = "pg_catalog",
  alias = "lake"
)
```

## Command-Line utilities

CLI tools are provided for VCF to Parquet conversion and querying, including threaded chunking based on contigs using either the [`bcf_reader`](./inst/scripts/vcf2parquet_duckdb.R) extension or via [Arrow IPC](./inst/scripts/vcf2parquet.R) 

```{bash cli-tool, eval=TRUE}
# Get paths using system.file
SCRIPT=$(Rscript -e "cat(system.file('scripts', 'vcf2parquet_duckdb.R', package='RBCFTools'))")
BCF=$(Rscript -e "cat(system.file('extdata', 'test_deep_variant.vcf.gz', package='RBCFTools'))")
OUT_PQ=$(mktemp --suffix=.parquet)
Log=$(mktemp --suffix=.log)

# Convert BCF to Parquet
time $SCRIPT convert --quiet --tidy -i $BCF -o $OUT_PQ -t 4  > $Log 2>&1

cat $Log

# Query with DuckDB SQL
$SCRIPT query -i $OUT_PQ -q "SELECT * FROM parquet_scan('$OUT_PQ') LIMIT 5"

# Describe table structure
$SCRIPT query -i $OUT_PQ -q "DESCRIBE SELECT * FROM parquet_scan('$OUT_PQ')"

# Show schema
$SCRIPT schema  --quiet -i $BCF 

# File info
$SCRIPT info -i $OUT_PQ 

rm -f $OUT_PQ
```



## References

- [Samtools documentation](https://www.htslib.org/doc/samtools.html)

- [Samtools GitHub](https://github.com/samtools/samtools)

- [FastDup GitHub](https://github.com/zzhofict/FastDup)

- [FastDup overlay errata](ERRATA.md)

- [BCFtools documentation](https://samtools.github.io/bcftools/)

- [BCFtools GitHub](https://github.com/samtools/bcftools)

- [HTSlib GitHub](https://github.com/samtools/htslib)

- [arrow-nanoarrow](https://arrow.apache.org/nanoarrow/)

- [Ducklake minio Example](https://github.com/duckdb/ducklake/blob/main/examples/minio-demo-server/README.md)
