---
output:
  github_document:
    toc: false
    html_preview: false
---

<!-- Generated as README.md from README.Rmd. Edit README.Rmd or ../run-short-read-pipelines.R, then rerun the benchmark. -->

# Complete HG02088 exome executable-pipeline benchmark

This is an executed full-input benchmark of the generic RBCFTools process
pipeline, not a synthetic package fixture. It maps all 10,110,535 read pairs
from the HG02088 exome run with BWA and Rminibwa, prepares one valid
name-sort/`fixmate -m`/coordinate-sort BAM per mapper, and runs both Samtools
`markdup` and the RBCFTools-patched FastDup against each prepared BAM.

Large references, FASTQs, indexes, and BAMs remain outside Git. [`../run-short-read-pipelines.R`](../run-short-read-pipelines.R) records
results incrementally, validates every BAM and index, records duplicate-QNAME
digests, retains text logs/metrics, and removes bulk outputs after successful
validation.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
performance <- read.csv("performance.csv", stringsAsFactors = FALSE, check.names = FALSE)
validation <- read.csv("validation.csv", stringsAsFactors = FALSE, check.names = FALSE)
environment <- readLines("environment.txt")
environment <- strsplit(environment, "=", fixed = TRUE)
environment <- setNames(
  vapply(environment, function(x) paste(x[-1L], collapse = "="), character(1L)),
  vapply(environment, `[[`, character(1L), 1L)
)
```

## Inputs and executable authorities

The human reference is the 1000 Genomes hs37d5 FASTA:

- <https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/reference/phase2_reference_assembly_sequence/hs37d5.fa.gz>
- 86 sequences and 3,137,454,505 unpadded bases.

The reads are the complete real 1000 Genomes phase 3 HG02088 exome run
SRR716421, classified as exome by the project sequence index:

- mate 1: <https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/HG02088/sequence_read/SRR716421_1.filt.fastq.gz>
- mate 2: <https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/HG02088/sequence_read/SRR716421_2.filt.fastq.gz>
- workload: 10,110,535 records in each mate (20,221,070 input reads).

Stable locators and workload roles are also recorded in
[`input-manifest.csv`](input-manifest.csv).

```{r environment-table}
knitr::kable(
  data.frame(
    Field = c(
      "Run UTC", "Host", "Platform", "CPU", "Logical CPUs", "Memory bytes",
      "Kernel", "RBCFTools", "RBCFTools commit", "Source tree state", "FastDup", "Samtools", "Rminibwa",
      "Pinned logical CPUs", "Physical-core sibling sets", "Pinned CPU max frequencies (kHz)",
      "Mapper/marker threads",
      "Sort threads per sort", "Fixmate threads", "Marker repetitions", "Cache policy",
      "Reference bytes", "Mate 1 bytes", "Mate 2 bytes"
    ),
    Value = c(
      environment[["run_utc"]], environment[["host"]], environment[["platform"]],
      environment[["cpu_model"]], environment[["logical_cpus"]],
      format(as.numeric(environment[["memory_bytes"]]), big.mark = ",", scientific = FALSE),
      environment[["kernel"]], environment[["rbcftools_version"]],
      environment[["rbcftools_commit"]], environment[["rbcftools_tree_state"]],
      environment[["fastdup_version"]],
      environment[["samtools_version"]], environment[["minibwa_version"]],
      environment[["cpu_affinity"]], environment[["cpu_thread_siblings"]],
      environment[["cpu_max_frequency_khz"]], environment[["threads"]], environment[["sort_threads"]], environment[["fixmate_threads"]],
      environment[["marker_repetitions"]], environment[["cache_policy"]],
      format(as.numeric(environment[["reference_bytes"]]), big.mark = ",", scientific = FALSE),
      format(as.numeric(environment[["read1_bytes"]]), big.mark = ",", scientific = FALSE),
      format(as.numeric(environment[["read2_bytes"]]), big.mark = ",", scientific = FALSE)
    )
  ),
  align = c("l", "r")
)
```

## Executed method

For each mapper, the benchmark builds a full hs37d5 index or explicitly records reuse of a prebuilt index. The reported BWA index was previously produced once with `bwa index -p <cache>/bwa-hs37d5 hs37d5.fa`; the runner hard-linked its five sidecars into the run directory. Remove `RBCFTOOLS_BENCH_BWA_INDEX_CACHE` from the command below to time a fresh BWA index build instead.

The mapping and preparation phase then executes:

```text
BWA mem or minibwa map
  | samtools sort -n
  | samtools fixmate -m
  | samtools sort (coordinate)
  > prepared.bam
```

That one prepared BAM is then consumed independently by:

```text
samtools markdup prepared.bam samtools.bam
samtools index samtools.bam

fastdup --input prepared.bam --output fastdup.bam ...
samtools index fastdup.bam
```

This isolates duplicate-marker time from mapping and sort time while also
exercising the complete executable pipeline. FastDup receives a seekable file
because its algorithm rereads coordinate-sorted input; it is not represented
as a streaming stage.

Every timed process is confined to the same explicit logical-CPU set with
`run_pipeline(cpu_affinity = ...)`. The selected IDs contain one SMT sibling
from each physical core; the runner rejects duplicate `thread_siblings_list`
values. All concurrently active stages share that one CPU set, so pipeline
stages cannot silently multiply the benchmark's physical-core budget. Internal
thread flags are still recorded because they describe scheduling demand, not
additional entitled cores.

RBCFTools samples each live stage's Linux `/proc/<pid>/status` every 10 ms.
`pipeline_peak_rss_kib` is the largest simultaneous sum of stage RSS values;
it is aggregate RSS and may double-count shared pages, rather than unique PSS.
`pipeline_peak_threads` is the largest simultaneous sum of live stage thread
counts. These are pipeline-level high-water marks; GNU `time -v` around only
one sibling process would not measure them.

## Measured wall time

```{r performance-table}
shown <- performance
shown$wall_seconds <- ifelse(is.na(shown$wall_seconds), "", sprintf("%.3f", shown$wall_seconds))
shown$peak_rss_kib <- ifelse(
  is.na(shown$peak_rss_kib),
  "",
  sprintf("%.3f", shown$peak_rss_kib / 1024^2)
)
shown$output_bytes <- ifelse(
  is.na(shown$output_bytes),
  "",
  format(shown$output_bytes, big.mark = ",", scientific = FALSE)
)
names(shown) <- c(
  "Mapper", "Phase", "Tool", "Repetition", "Cache", "Wall seconds",
  "Peak aggregate RSS (GiB)", "Peak simultaneous threads", "Pinned CPUs",
  "Internal thread configuration", "Stage statuses", "Output bytes"
)
knitr::kable(shown, align = c("l", "l", "l", "r", "l", "r", "r", "r", "l", "l", "l", "r"))
```

```{r marker-comparison}
marker_rows <- subset(performance, phase == "duplicate marking")
marker_summary <- do.call(
  rbind,
  lapply(split(marker_rows, list(marker_rows$mapper, marker_rows$tool), drop = TRUE), function(x) {
    data.frame(
      Mapper = x$mapper[[1L]],
      Tool = x$tool[[1L]],
      `Minimum wall (s)` = min(x$wall_seconds),
      `Median wall (s)` = median(x$wall_seconds),
      `Maximum wall (s)` = max(x$wall_seconds),
      `Median peak RSS (GiB)` = median(x$peak_rss_kib) / 1024^2,
      `Median peak threads` = median(x$peak_threads),
      check.names = FALSE
    )
  })
)
marker_summary[3:7] <- lapply(marker_summary[3:7], function(x) sprintf("%.3f", x))
row.names(marker_summary) <- NULL
knitr::kable(marker_summary, align = c("l", "l", "r", "r", "r", "r", "r"))
```

Duplicate-marker runs use three repetitions with alternating tool order and a
cold page cache before each timed phase. The table establishes what was
observed for this complete exome on one pinned host; it does not claim a
general speedup outside the declared workload and resource contract.

## Output validation

Every output passed `samtools quickcheck`, was indexed by the bundled Samtools,
and was scanned to count all records and duplicate-flagged records. Duplicate
QNAMEs were deduplicated, sorted, and hashed so marker agreement is auditable
without retaining multi-gigabyte BAMs.

```{r validation-table}
shown_validation <- validation
shown_validation$bam_bytes <- format(shown_validation$bam_bytes, big.mark = ",", scientific = FALSE)
shown_validation$bai_bytes <- format(shown_validation$bai_bytes, big.mark = ",", scientific = FALSE)
names(shown_validation) <- c(
  "Mapper", "Marker", "Repetition", "Records", "Duplicate records", "Duplicate QNAMEs",
  "Duplicate-QNAME SHA-256", "BAM bytes", "BAI bytes", "Quickcheck"
)
knitr::kable(shown_validation, align = c("l", "l", "r", "r", "r", "r", "l", "r", "r", "l"))
```

```{r duplicate-agreement}
samtools_validation <- subset(validation, marker == "samtools")
fastdup_validation <- subset(validation, marker == "fastdup")
agreement <- merge(
  samtools_validation,
  fastdup_validation,
  by = c("mapper", "repetition"),
  suffixes = c(".samtools", ".fastdup")
)
agreement <- data.frame(
  Mapper = agreement$mapper,
  Repetition = agreement$repetition,
  `Same duplicate-record count` = agreement$duplicate_records.samtools == agreement$duplicate_records.fastdup,
  `Same duplicate-QNAME count` = agreement$duplicate_qnames.samtools == agreement$duplicate_qnames.fastdup,
  `Same duplicate-QNAME digest` = agreement$duplicate_qname_sha256.samtools == agreement$duplicate_qname_sha256.fastdup,
  check.names = FALSE
)
knitr::kable(agreement, align = c("l", "r", "l", "l", "l"))
```

Matching duplicate counts alone do not prove identical decisions; matching the
sorted unique-QNAME digest provides a stronger marker-level comparison. Any
intentional differences caused by the RBCFTools FastDup corrections remain
governed by the package root `ERRATA.md`.
