---
title: "BCF Reader Benchmarks"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
params:
  bcf_path: "test_very_large.bcf"
  region: "22:20000000-21000000"
  duckdb_ext: "build/bcf_reader.duckdb_extension"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

Quick, reproducible timing comparisons between:

- `bcftools view` streamed to `/dev/null`
- DuckDB `bcf_read()` via the packaged extension

The large input (`test_very_large.bcf`) is expected to live locally (never committed); generated from GLIMPSE-imputed 1000 Genomes data.

## Prerequisites

- `bcftools` available on `PATH`
- DuckDB CLI installed and `duckdb` R package (for rendering)
- Built extension at `params$duckdb_ext` (`make` in this folder)
- Large BCF at `params$bcf_path` (defaults to `test_very_large.bcf` in this directory)

```{r paths}
bcf_path <- params$bcf_path
region <- params$region
duckdb_ext <- params$duckdb_ext

if (!file.exists(bcf_path)) {
  stop("BCF not found: ", bcf_path, ". Drop the file locally before running benchmarks.")
}
if (!file.exists(duckdb_ext)) {
  stop("DuckDB extension not found: ", duckdb_ext, ". Build with `make` first.")
}
```

## Benchmarks

Timings are single runs using `system.time()` for wall-clock comparisons; adjust or repeat as needed. Region queries mirror full-file timings but with `-r` / `region :=`.

```{r helpers}
run_cmd <- function(cmd) {
  system.time({
    out <- system(cmd, intern = TRUE, ignore.stdout = TRUE, ignore.stderr = FALSE)
    attr(out, "status") <- attr(out, "status")
    out
  })
}
```

### bcftools view (full file)

```{r bcftools-full}
cmd_bcftools_full <- sprintf("bcftools view %s > /dev/null", shQuote(bcf_path))
time_bcftools_full <- run_cmd(cmd_bcftools_full)
time_bcftools_full
```

### bcftools view (region)

```{r bcftools-region}
cmd_bcftools_region <- sprintf("bcftools view -r %s %s > /dev/null", shQuote(region), shQuote(bcf_path))
time_bcftools_region <- run_cmd(cmd_bcftools_region)
time_bcftools_region
```

### DuckDB bcf_read (full file)

```{r duckdb-full}
cmd_duckdb_full <- sprintf(
  "duckdb -unsigned -c \"LOAD %s; SELECT COUNT(*) FROM bcf_read(%s);\"",
  shQuote(duckdb_ext),
  shQuote(bcf_path)
)
time_duckdb_full <- run_cmd(cmd_duckdb_full)
time_duckdb_full
```

### DuckDB bcf_read (region)

```{r duckdb-region}
cmd_duckdb_region <- sprintf(
  "duckdb -unsigned -c \"LOAD %s; SELECT COUNT(*) FROM bcf_read(%s, region := %s);\"",
  shQuote(duckdb_ext),
  shQuote(bcf_path),
  shQuote(region)
)
time_duckdb_region <- run_cmd(cmd_duckdb_region)
time_duckdb_region
```

## Notes

- These timings are single passes; rerun and average for more stable numbers.
- Ensure the file and its index are co-located with fast storage (local SSD preferred).
- Adjust `region` to match indexed contigs present in `test_very_large.bcf`.
