---
title: "BCF Reader Benchmarks"
output: github_document
params:
  bcf_path: "../../test_very_large.bcf"
  region: "22:20000000-21000000"
  duckdb_ext: "build/bcf_reader.duckdb_extension"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

Quick, reproducible timing comparisons between:

- `bcftools view` streamed to `/dev/null`
- DuckDB `bcf_read()` via the packaged extension

The large input (`test_very_large.bcf`) is expected to live locally (never committed); generated from GLIMPSE-imputed 1000 Genomes data. Each command is run twice by default to surface “cold-ish” vs “warm-ish” cache timing (adjust `n_runs` below).

## Prerequisites

- `bcftools` available on `PATH`
- DuckDB CLI installed and `duckdb` R package (for rendering)
- Built extension at `params$duckdb_ext` (`make` in this folder)
- Large BCF at `params$bcf_path` (defaults to `test_very_large.bcf` in this directory)

```{r paths}
bcf_path <- params$bcf_path
region <- params$region
duckdb_ext <- params$duckdb_ext

has_bcf <- file.exists(bcf_path)
has_ext <- file.exists(duckdb_ext)
can_run <- has_bcf && has_ext

if (!has_bcf) message("BCF not found: ", bcf_path, ". Benchmarks will be skipped.")
if (!has_ext) message("DuckDB extension not found: ", duckdb_ext, ". Build with `make` first; benchmarks will be skipped.")

data.frame(
  resource = c("bcf_path", "duckdb_ext"),
  path = c(bcf_path, duckdb_ext),
  exists = c(has_bcf, has_ext)
)
```

## Benchmarks

Timings are single runs using `system.time()` for wall-clock comparisons; adjust or repeat as needed. Region queries mirror full-file timings but with `-r` / `region :=`.

```{r helpers}
run_cmd <- function(cmd) {
  t <- system.time({
    system(cmd, intern = TRUE, ignore.stdout = TRUE, ignore.stderr = FALSE)
  })
  c(elapsed = unname(t["elapsed"]),
    user = unname(t["user.self"]),
    system = unname(t["sys.self"]))
}

n_runs <- 2  # adjust to change cold/warm repetition

time_runs <- function(cmd, runs = n_runs) {
  if (!can_run) return(NULL)
  times <- t(vapply(seq_len(runs), function(i) run_cmd(cmd), numeric(3)))
  data.frame(
    run = seq_len(runs),
    elapsed = times[, "elapsed"],
    user = times[, "user"],
    system = times[, "system"],
    cache_state = ifelse(seq_len(runs) == 1, "cold-ish", "warm-ish")
  )
}
```

### bcftools view (full file)

```{r bcftools-full, eval=can_run}
cmd_bcftools_full <- sprintf("bcftools view %s > /dev/null", shQuote(bcf_path))
res_bcftools_full <- time_runs(cmd_bcftools_full)
res_bcftools_full
```

### bcftools view (region)

```{r bcftools-region, eval=can_run}
cmd_bcftools_region <- sprintf("bcftools view -r %s %s > /dev/null", shQuote(region), shQuote(bcf_path))
res_bcftools_region <- time_runs(cmd_bcftools_region)
res_bcftools_region
```

### DuckDB bcf_read (full file)

```{r duckdb-full, eval=can_run}
cmd_duckdb_full <- sprintf(
  "duckdb -unsigned -c \"LOAD %s; SELECT COUNT(*) FROM bcf_read(%s);\"",
  shQuote(duckdb_ext),
  shQuote(bcf_path)
)
res_duckdb_full <- time_runs(cmd_duckdb_full)
res_duckdb_full
```

### DuckDB bcf_read (region)

```{r duckdb-region, eval=can_run}
cmd_duckdb_region <- sprintf(
  "duckdb -unsigned -c \"LOAD %s; SELECT COUNT(*) FROM bcf_read(%s, region := %s);\"",
  shQuote(duckdb_ext),
  shQuote(bcf_path),
  shQuote(region)
)
res_duckdb_region <- time_runs(cmd_duckdb_region)
res_duckdb_region
```

## Notes

- These timings are single passes; rerun and average for more stable numbers.
- Ensure the file and its index are co-located with fast storage (local SSD preferred).
- Adjust `region` to match indexed contigs present in `test_very_large.bcf`.
- If required resources are missing, chunks are skipped and a status table is shown above.
